A note on the instability and degeneracy of deep learning models

### 2017-05-10 writing/tech_submission/appendix/appendix-note.Rmd
forgot to add old files

{\bf Proof of Proposition \ref{prop:instab}.} We prove the contrapositive, supposing that $\Delta(\boldsymbol \theta_N) \le C$ holds for some $C > 0$ and show $\text{ELPR}(\boldsymbol \theta_N) \leq NC$. Let $\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$ and $\boldsymbol x_{max} \equiv \argmax\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$. Note there exists a sequence $\boldsymbol x_{min} \equiv \boldsymbol x_0, \boldsymbol x_1, \dots, \boldsymbol x_k \equiv \boldsymbol x_{max}$ in $\mathcal{X}^N$ of component-wise switches to move from $\boldsymbol x_{min}$ to $\boldsymbol x_{max}$ in the sample space (i.e. $\boldsymbol x_i, \boldsymbol x_{i + 1} \in \mathcal{X}^N$ differ in exactly $1$ component for $i = 0, \dots, k$) for some integer $k \in \{0, 1, \dots, N\}$. Under the FSFS model, recall $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ holds so that $\log P_{\boldsymbol \theta_N}(\boldsymbol x)$ is well-defined for each outcome $\boldsymbol x \in \mathcal{X}^N$. Then, if $k > 0$, it follows that
$$
\text{ELPR}(\boldsymbol \theta_N) = \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] = \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_{i-1})}\right)\right| \le \sum\limits_{i = 1}^k\left|\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta}(\boldsymbol x_{i-1})}\right)\right| \le k \Delta_N(\boldsymbol \theta_N) \le NC,
$$
using $k \le N$ and $\Delta(\boldsymbol \theta_N) \le C$. If $k = 0$, then $\boldsymbol x_{max} = \boldsymbol x_{min}$ and the same bound above holds.$\Box$

>>>>

{\bf Proof of Proposition \ref{prop:instab}.} We prove the contrapositive, supposing that $\Delta(\boldsymbol \theta_N) \le C$ holds for some $C > 0$ and show $\text{ELPR}(\boldsymbol \theta_N) \leq NC$. Let $\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$ and $\boldsymbol x_{max} \equiv \argmax\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$. Note there exists a sequence $\boldsymbol x_{min} \equiv \boldsymbol x_0, \boldsymbol x_1, \dots, \boldsymbol x_k \equiv \boldsymbol x_{max}$ in $\mathcal{X}^N$ of component-wise switches to move from $\boldsymbol x_{min}$ to $\boldsymbol x_{max}$ in the sample space (i.e. $\boldsymbol x_i, \boldsymbol x_{i + 1} \in \mathcal{X}^N$ differ in exactly $1$ component for $i = 0, \dots, k$) for some integer $k \in \{0, 1, \dots, N\}$. Under the FSFS model, recall $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ holds so that $\log P_{\boldsymbol \theta_N}(\boldsymbol x)$ is well-defined for each outcome $\boldsymbol x \in \mathcal{X}^N$. Then, if $k > 0$, it follows that
\begin{align*}
\text{ELPR}(\boldsymbol \theta_N) = \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] &= \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_{i-1})}\right)\right| \\
&\le \sum\limits_{i = 1}^k\left|\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta}(\boldsymbol x_{i-1})}\right)\right| \le k \Delta_N(\boldsymbol \theta_N) \le NC,
\end{align*}
using $k \le N$ and $\Delta(\boldsymbol \theta_N) \le C$. If $k = 0$, then $\boldsymbol x_{max} = \boldsymbol x_{min}$ and the same bound above holds.$\Box$

### 2017-05-10 writing/tech_submission/note.Rmd
forgot to add old files

---
title: "A note on the instability and degeneracy of deep learning models"
author: 
  - name: Andee Kaplan
    university: Iowa State University
    department: Department of Statistics
  - name: Daniel Nordman
    university: Iowa State University
    department: Department of Statistics
  - name: Stephen Vardeman
    university: Iowa State University
    department: |
      | Department of Statistics and Industrial Manufacturing 
      | Systems Engineering Department
output: 
  pdf_document:
    keep_tex: false
    fig_caption: yes
    includes:
      in_header: latex/header.tex
    number_sections: yes
    template: latex/technometrics_article.tex
fontsize: 12pt
bibliography: references/refs.bib
abstract: |
  A probability model exhibits instability if small changes in a data outcome result in large, and often unanticipated, changes in probability. For correlated data structures found in several application areas, there is increasing interest in predicting/identifying instability. We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of length $N$ has a finite number of possible outcomes. (A sequence of probability models results indexed by $N$ that accommodates data of expanding dimension.) Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than $N$. In this case, a one component change in the data sequence can shift probability by orders of magnitude. Also, as a measure of instability becomes more extreme, the resulting probability models are shown  to tend to degeneracy, placing all their probability on arbitrarily small portions of the sample space. These results on instability apply to large classes of models commonly used in random graphs, network analysis, and machine learning contexts.
keywords: Degeneracy, Instability, Deep Learning, Graphical Models
blind: false
---

>>>>

---
title: "A note on the instability and degeneracy of deep learning models"
author: 
  - name: Andee Kaplan
    university: Iowa State University
    department: Department of Statistics
  - name: Daniel Nordman
    university: Iowa State University
    department: Department of Statistics
  - name: Stephen Vardeman
    university: Iowa State University
    department: |
      | Department of Statistics and Department of Industrial and
      | Manufacturing Systems Engineering
output: 
  pdf_document:
    keep_tex: true
    fig_caption: yes
    includes:
      in_header: latex/header.tex
    number_sections: yes
    template: latex/technometrics_article.tex
fontsize: 12pt
bibliography: references/refs.bib
abstract: |
  A probability model exhibits instability if small changes in a data outcome result in large, and often unanticipated, changes in probability. For correlated data structures found in several application areas, there is increasing interest in predicting/identifying instability. We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of length $N$ has a finite number of possible outcomes. (A sequence of probability models results indexed by $N$ that accommodates data of expanding dimension.) Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than $N$. In this case, a one component change in the data sequence can shift probability by orders of magnitude. Also, as a measure of instability becomes more extreme, the resulting probability models are shown  to tend to degeneracy, placing all their probability on arbitrarily small portions of the sample space. These results on instability apply to large classes of models commonly used in random graphs, network analysis, and machine learning contexts.
keywords: Degeneracy, Instability, Deep Learning, Graphical Models
blind: false
---
